---
title: "Air Quality Time Series Report"
author: "Adam Sampson, Jay Bektasevic, Brenden McGale, Andrew Brill, Taylor Maxson"
date: "February 20, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
# List of packages to load:
packages <- c("tidyverse", "lubridate", "tibbletime", "rlang", "dygraphs", "forecast", "zoo", "xts", "stringr", "knitr", "kableExtra", "ggplot2", "tseries")
  
# Check to see whether any packages aren't installed on the computer and install
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)
  
# Load Neccessary Packages
sapply(packages, require, character.only = TRUE)
rm(new_packages)
```

# **Introduction**

This report is an analysis of data related to air quality. The exploratory question this report investigates is: How closely does weather data correlate with air quality data in Louisville?

# **Datasets**

## Weather Data

Hourly weather data was retrieved for Bowman Field Weather Station using the NCDC / NOAA simplified weather data download form at: <https://www7.ncdc.noaa.gov/CDO/cdopoemain.cmd?datasetabbv=DS3505&countryabbv=&georegionabbv=&resolution=40>.

```{r echo=FALSE}
load('weather_info.RData')
weather_desc.df <- 
  data.frame(matrix(names(weather.info$Description),
                    nrow = 33)) %>%
  data.frame(matrix(unlist(weather.info$Description),
                    nrow = 33, 
                    ncol = 1, 
                    byrow = T)) 
names(weather_desc.df) <- c("col_name","description")
weather_desc.df <- weather_desc.df %>%
  filter(col_name == "DATE_TIME"|
           col_name == "DIR"|
           col_name == "SPD"|
           col_name == "TEMP"|
           col_name == "PCP01")
rm(weather.info)
knitr::kable(weather_desc.df, booktabs = TRUE) 
```

## Air Quality Data

Particulate matter (PM~2.5~) is a broad term used for an airborne mixture of solid particles and liquid droplets. These particles are <= 2.5 microns in diameter and, although the composition is region-specific, are largely made up of sulphate, nitrate, carbon particles and soil. Fine particles are produced from all types of combustion, including motor vehicles, power plants, residential wood burning, forest fires, agricultural burning, and some industrial processes.
 
 
Along with ground-level ozone, fine particulate matter is one of the two major components of smog.
 
From a human perspective, excess fine particulates pose hazards for people with asthma, cardiovascular or lung disease, as well as children and the elderly. These health effects have been associated with both short term (daily) and long term (>year) exposure. However, even if you are healthy, you may feel temporary symptoms if you are exposed to high levels of particle pollution. Numerous scientific studies connect particle pollution exposure to a variety of health issues, including:
    
*   irritation of the eyes, nose and throat
*   coughing, chest tightness and shortness of breath
*   reduced lung function
*   irregular heartbeat
*   asthma attacks
*   heart attacks
*   premature death in people with heart or lung disease

Ecologically, fine particulate matter can damage vegetation and can lead to soil erosion.

The EPA has set a National Ambient Air Quality Standards for six principal pollutants found [here](https://www.epa.gov/criteria-air-pollutants/naaqs-table).

Air Quality Index Table

| AQ Index  | Health Concern | Meaning                                      |
|-----------|----------------|----------------------------------------------|
|0 to 50	|Good   | Air quality is considered satisfactory, and air pollution poses little or no risk.|
|51 to 100 |Moderate| Air quality is acceptable; however, for some pollutants there may be a moderate health concern for a very small number of people who are unusually sensitive to air pollution.|
|101 to 150	|Unhealthy for Sensitive Groups| Members of sensitive groups may experience health effects. The general public is not likely to be affected.|	
|151 to 200	|Unhealthy| Everyone may begin to experience health effects; members of sensitive groups may experience more serious health effects.|
|201 to 300	|Very Unhealthy| Health alert: everyone may experience more serious health effects.|
|301 to 500	|Hazardous| Health warnings of emergency conditions. The entire population is more likely to be affected.|

Data was aquired from the EPA [website](https://www.epa.gov/outdoor-air-quality-data/download-daily-data) there are several monitoring stations in Jefferson County.
```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
#   Gather all the files in the directory
    temp <- list.files(pattern="*.csv")

#  Read in all the files and bind them together
   EPA_data <- lapply(temp, function(x) read.csv(x, stringsAsFactors = FALSE)) %>% 
               bind_rows()
   
   save(EPA_data, file = "EPA.Rdata")

```

### Exploratory Analysis of Air Data
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
    
    load("EPA_data.Rdata")
    #EPA_data <- read_excel("EPA_data.xlsx")

#   Choose two years to compare. Also, this will average the measurements accross all sites.

    ozone <- EPA_data[EPA_data$AQS_PARAMETER_DESC =="Ozone", ] %>% 
            group_by(Date) %>% 
            summarise(avg.measurement = mean( Measurement )) %>% 
            mutate(ozone_7 = rollmean(avg.measurement, k = 7, fill = NA),
                   ozone_30 = rollmean(avg.measurement, k = 30, fill = NA),
                   ozone_60 = rollmean(avg.measurement, k = 60, fill = NA)
                   )                   
    so2 <- EPA_data[EPA_data$AQS_PARAMETER_DESC =="Sulfur dioxide", ] %>% 
            group_by(Date) %>% 
            summarise(avg.measurement = mean( Measurement )) %>% 
            mutate(so2_7 = rollmean(avg.measurement, k = 7, fill = NA),
                   so2_30 = rollmean(avg.measurement, k = 30, fill = NA),
                   so2_60 = rollmean(avg.measurement, k = 60, fill = NA)
            )
    
    pm2.5 <- EPA_data[EPA_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", ]%>% 
            group_by(Date) %>% 
            summarise(avg.measurement = mean( Measurement )) %>% 
            mutate(pm2.5_7 = rollmean(avg.measurement, k = 7, fill = NA),
                   pm2.5_30 = rollmean(avg.measurement, k = 30, fill = NA),
                   pm2.5_60 = rollmean(avg.measurement, k = 60, fill = NA)
            )
    
    AQI <- EPA_data %>% select(Date, DAILY_AQI_VALUE) %>% 
            group_by(Date) %>% 
            summarise(avg.measurement = mean( DAILY_AQI_VALUE )) %>% 
            mutate(AQI_7 = rollmean(avg.measurement, k = 7, fill = NA),
                   AQI_30 = rollmean(avg.measurement, k = 30, fill = NA),
                   AQI_60 = rollmean(avg.measurement, k = 60, fill = NA)
            )
    

```

```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, paged.print=FALSE}


    par(mfcol = c(2, 2)  )
    plot(ozone$Date , ozone$avg.measurement ,  xlim = range(ozone$Date, na.rm = TRUE), 
         ylim = range(ozone$avg.measurement, na.rm = T), xlab = "", ylab = expression(O[3] * " (ppm)"))
    title("Ozone (1990 - 2017)")
    # Need to figure out to plot the moveing average
    lines( ozone$Date  ,  ozone$ozone_60 , type="l", pch=50, lty=2, col="blue")  
    abline(h = 0.070, col = "red")
    # legend("topright", legend=c("EPA guideline ", "MA 60 days"),
    #    col=c("red", "blue"), lty=1:1, cex=0.8,
    #    box.lty=0)
    
    plot(so2$Date , so2$avg.measurement ,  xlim = range(so2$Date, na.rm = TRUE), 
         ylim = range(so2$avg.measurement, na.rm = T), xlab = "", ylab = expression(SO[2] * " (ppm)"))
    title("Sulfur Dioxide (1990 - 2017)")
     lines( so2$Date  ,  so2$so2_60 , type="l", pch=50, lty=2, col="blue")
      # legend("topright", legend=c("EPA guideline ", "MA 60 days"),
      #   col=c("red", "blue"), lty=1:1, cex=0.8,
      #   box.lty=0)
    abline(h = 75, col = "red")
    
    #par(mfrow = c(1,2))
    plot(pm2.5$Date,pm2.5$avg.measurement , xlim = range(pm2.5$Date, na.rm = TRUE), 
         ylim = range(pm2.5$avg.measurement, na.rm = T), xlab = "", 
         ylab = expression(PM[2.5] * " (" * mu * g/m^3 * ")"))
    title(expression(PM[2.5] * " (2000 - 2017)"))
     lines( pm2.5$Date  ,  pm2.5$pm2.5_60 , type="l", pch=50, lty=2, col="blue")
     legend("topright", legend=c("EPA guideline ", "MA 60 days"),
        col=c("red", "blue"), lty=1:1, cex=0.8,
        box.lty=0)
    abline(h = 12.0, col = "red") 
    
     plot(AQI$Date,AQI$avg.measurement ,  xlim = range(AQI$Date, na.rm = TRUE), 
         ylim = range(AQI$avg.measurement, na.rm = T), xlab = "", 
         ylab = "Air Quality Index ")
    title("Air Quality Index (1990 - 2017)")
        lines( AQI$Date  ,  AQI$AQI_60 , type="l", pch=50, lty=2, col="blue")
     # legend("topright", legend=c("EPA guideline ", "MA 60 days"),
     #   col=c("red", "blue"), lty=1:1, cex=0.8,
     #   box.lty=0)
    abline(h = 50, col = "red")
```

### Seasonal Variation in Air Quality
```{r  fig.align='center' ,echo=FALSE}
    par(mfrow = c(2, 1))
    pm2.5_ts <- ts(pm2.5$avg.measurement, start = c(2000, 1), end = c(2017,09),  frequency = 12)
    plot(pm2.5_ts,  ylab = expression(PM[2.5] * " (" * mu * g/m^3 * ")"))
    title("PM 2.5 2000 - 2017")
    
    seasonplot(pm2.5_ts, s = 12,   main=expression(PM[2.5] * " Seasonal Plot"),
                year.labels=TRUE, year.labels.left=FALSE, ylab = expression(PM[2.5] * " (" * mu * g/m^3 * ")"), xlab = "", col=1:20, pch=1)
    
    par(mfrow = c(2,1))
    decomp_pm <- decompose(pm2.5_ts, type = "multiplicative" )
    acf(pm2.5_ts)
    acf(diff(pm2.5_ts, lag = 5))
    adf.test(pm2.5_ts, alternative = "stationary")
    plot(decomp_pm)
 
```

For a stationary time series, the ACF will drop to zero relatively quickly, while the ACF of non-stationary data decreases slowly.


### Correlations with other pollutants
In order to examine the cause of pollution we will need to look at correlation between the PM~2.5~ and other weather variables such as temperature, wind, and precipitation.  


```{r echo=FALSE, message=FALSE, warning=FALSE}
    load(file = "BowmanField_Weather_daily.RData")
#   Convert date to common format for merging
    bf.daily.tt$DATE_TIME <- as.Date(bf.daily.tt$DATE_TIME, format = "%Y-%m-%d")
    EPA_data$Date <- as.Date(EPA_data$Date, format = "%Y-%m-%d")
#   AQI$Date <- as.Date(AQI$Date, format = "%Y-%m-%d")
#   AQI_merge <- as.data.frame( left_join(AQI, bf.daily.tt, by = c("Date"="DATE_TIME")))
#   Merge the two datasets
    comb_data <- as.data.frame( left_join(EPA_data, bf.daily.tt, by = c("Date"="DATE_TIME")))
```

Now we can plot the variables and see if they corrlate with each other.
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
        par(mfrow = c(2, 2))
         plot(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "mean.temp"], 
             comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"],
             xlab = "Mean Temp. (F)", ylab = expression(PM[2.5] * " (" * mu * g/m^3 * ")"))
        title(expression(PM[2.5] * " vs. Mean Temp. (1990 - 2017)"))
        abline(lm(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"] ~
                  comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "mean.temp"]), col = "red")
        # print(cor(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "mean.temp"], 
        #      comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"]))


        
           plot(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "mean.spd"], 
             comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"],
             xlab = "Mean Wind Speed", ylab = expression(PM[2.5] * " (" * mu * g/m^3 * ")"))
        title(expression(PM[2.5] * " vs. Mean Wind Spd. (1990 - 2017)"))
        abline(lm(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"] ~
                  comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "mean.spd"]), col = "red")
        # print(cor(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "mean.spd"], 
        #      comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"]))
                
        
          plot(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "sum.precip"], 
             comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"],
             xlab = "Precipitation (Inches)", ylab = expression(PM[2.5] * " (" * mu * g/m^3 * ")"))
        title(expression(PM[2.5] * " vs. Precipitation (1990 - 2017)"))
        abline(lm(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"] ~
                  comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "sum.precip"]), col = "red")
        # print(cor(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "sum.precip"], 
        #      comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"]))
        
        
        plot(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "DAILY_AQI_VALUE"], 
             comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"],
             xlab = "Air Quality Index", ylab = expression(PM[2.5] * " (" * mu * g/m^3 * ")"))
        title(expression(PM[2.5] * " vs. AQI (1990 - 2017)"))
        abline(lm(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"] ~
                  comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "DAILY_AQI_VALUE"]), col = "red")
        # print(cor(comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "DAILY_AQI_VALUE"], 
        #      comb_data[comb_data$AQS_PARAMETER_DESC =="PM2.5 - Local Conditions", "Measurement"]))     
        

        
```

As shown above, there is a slight correlation between PM~2.5~ and mean temperature. It shows a slight increase in PM~2.5~ as the mean temperature increases.  Another interesting relationship can be observed by plotting PM~2.5~ vs. Mean Wind Speed –we can observe that the PM~2.5~ levels actually drop as the Mean wind speed increases. This is counter-intuitive because one would think that higher winds would cause PM~2.5~ levels to spike. This phenomenon can actually be explained, since, we are in the valley and all the pollution gets trapped in the valley, higher wind speeds are attributed to weather changes such as weather fronts that push out air from the valley - hence the lover readings. 

Increase in precipitation shows a similar relationship as with the wind speed. This phenomenon also makes sense as an increase in precipitation allows fine particulates to bind to rain droplets and fall to the ground. 

The obvious relationship is an increase in an Air Quality Index also proportionately increases PM~2.5~ levels. Note – PM~2.5~ levels are used in calculating AQI. 

### PM~2.5~ Forecast 

Now that we know that the time series is a stationary we can go ahead and forecast PM~2.5~ levels 12 months in the future. I will use Holt-Winters model as it does a great job of picking up seasonality and the trend.
```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
   
    pm2.5_forecast <- HoltWinters(pm2.5_ts)
    summary(forecast(pm2.5_forecast,12))
   checkresiduals(pm2.5_forecast)
    plot(forecast(pm2.5_forecast,12),  xlab="", ylab=expression(PM[2.5] * " (" * mu * g/m^3 * ")"), 
         main= expression(PM[2.5] * " Forecast (12 months)"), pch=25)
    abline(h = 12.0, col = "red") 
    legend("topleft", legend=c("Actual", "Forecast", "EPA guideline"),
       col=c("black", "blue", "red"), lty=1:1, cex=0.8,
       box.lty=0)
```

Even though the short-term forecast looks reasonable, this is a very crude model for a complex process. The residuals demonstrate that there is a lot of information that has not been captured with this model.

# **Techniques**

## Data Cleaning

### Cleaning Weather Data

| Variable  | Cleanup                       |
|-----------|-------------------------------|
| Date_TIME | Clean                         |
| DIR       | Any \* in field indicates NA  |
|           | 990 - Variable - NA           |
|           | \>360 - Uncertain - NA        |
| SPD       | Any \* in field indicates NA  |
| TEMP      | Any \* in field indicates NA  |
| PCP01     | Any \* in field indicates NA  |
  
## Converting Variables

### Converting Wind Compass Direction to Inverval Data

Wind data is recieved in compass degrees. This is problematic when performing calculations because 0/360 degrees is between 350 degrees and 10 degrees, but the mathematical average of 350 degrees and 10 degrees is 180 which is exactly the opposite of the true average direction.

One way to make this work better is to convert from polar coordinates to cartesian coordinates.

```{r fig.width=2, fig.height=2, fig.align='center' ,echo=FALSE}
gg_circle <- function(r, xc, yc, color="black", fill=NA, ...) {
    x <- xc + r*cos(seq(0, pi, length.out=100))
    ymax <- yc + r*sin(seq(0, pi, length.out=100))
    ymin <- yc + r*sin(seq(0, -pi, length.out=100))
    annotate("ribbon", x=x, ymin=ymin, ymax=ymax, color=color, fill=fill, ...)
}
  
ggplot() +
  xlim(-1.1,1.1) +
  ylim(-1.1,1.1) +
  gg_circle(r=1,xc=0,yc=0) +
  geom_text(aes(label = "0 deg"),x=0,y=1.1) +
  geom_text(aes(label = "180 deg"), x=0, y=-1.1) +
  geom_line(aes(x=c(0,0.707),y=c(0,0.707))) +
  geom_line(aes(x=c(0,0),y=c(0,0.707)),linetype = 2) +
  geom_line(aes(x=c(0,0.707),y=c(0.707,0.707)),linetype = 2) +
  geom_text(aes(label = "dir"), x=0.1,y=0.25) +
  theme_minimal() + 
  theme(aspect.ratio = 1,
        axis.title = element_blank(),
        axis.text = element_blank(),
        panel.background = element_blank(),
        panel.grid = element_blank())
```

$$ sin(dir) = \frac{x}{1} \rightarrow x = sin(dir*\frac{180}{\Pi}) $$
$$ cos(dir) = \frac{y}{1} \rightarrow y = cos(dir*\frac{180}{\Pi}) $$

An interesting bi-product of this transformation is that it is now possible to have a wind direction of {0,0} when there is no wind speed and lack of wind does not necessarily mean that wind direction has to be removed from calculations as NA.

## Aggregating Data

### Aggregating Weather Data to Daily

The Air Quality data is in daily format. Therefore it makes sense to aggregate the weather data to daily values as well. 

| Input     | Output(s)         |
|-----------|-------------------|
| TEMP      | min / mean / max  |  
| x.dir     | min / mean / max  |
| y.dir     | min / mean / max  |
| SPD       | min / mean / max  |
| PCP01     | sum               |

## Extracting addional information

### Extracting Frequency Data from Precipitation

## Dealing with NA values

### Imputing NA values in Weather Data

| Input     | Impute Method     |
|-----------|-------------------|
| TEMP      | na.approx         |  
| x.dir     | 0 if NA           |
| y.dir     | 0 if NA           |
| SPD       | na.approx         |
| PCP01     | 0 if NA           |

## Checking for Seasonality and Cycles

The Autocorrelation of variables was checked using the `acf()` function.

### Weather Data Seasonality

The charts below show that some of the weather data is seasonaly correlated over a year, and one of the sets of data is cyclically correlated to recent days.

* Wind Direction is not seasonally correlated. There is a minor correlation to the day before, but minimal correlation beyond 2 days.
* Wind speed is seasonally correlated on a yearly basis.
* Temperature is seasonally correlated on a yearly basis. 
  + As expected
* Daily Precipitation is autocorrelated all over the place. For Louisville it appears that there is often rain followed by several days of no rain and then repeating.
  + This is siginficantly different from western states where rain is seasonally correlated to the time of year with wet winters and dry summers.
  + It may be possible to correlate amount of rain on a seasonal basis still
* Days since last rain is autocorrelated over several days and falling off the farther you get from 0 days. This re-enforces the concept that rain is correlated heavily to whether it rained recently or not.

```{r fig.width=8, fig.height=8, echo=FALSE}
load('BowmanField_Weather_Daily.RData')

par(mfrow=c(3,2))
  acf(bf.daily.tt$mean.dir.x, na.action = na.pass, lag.max = 365, 
      main='Mean Wind Dir X')
  acf(bf.daily.tt$mean.dir.y, na.action = na.pass, lag.max = 365, 
      main='Mean Wind Dir Y')
  acf(bf.daily.tt$mean.spd, na.action = na.pass, lag.max = 365, 
      main='Mean Wind Speed')
  acf(bf.daily.tt$mean.temp, na.action = na.pass, lag.max = 365, 
      main='Mean Temperature (F)')
  acf(bf.daily.tt$sum.precip, na.action = na.pass, lag.max = 365, 
      main='Daily Precipitation')
  acf(bf.daily.tt$days.since.precip, na.action = na.pass, lag.max = 365, 
      main='Days Since Precipitation')
```

## Decomposition

### Weather Data Decomposition

For weather data it only makes sense to decompose Temperature, Wind Speed, and Rain Amount based on the results of the `acf()`. 

#### Mean Daily Temperature

The plots below show that mean daily temperature is seasonal over a yearly period. Interestingly, the random component is still seasonally affected. This is likely because the day-to-day variation changes between seasons. More advanced methods are needed to deal with this, however this decomposition should be sufficient for comparing to air quality data.

It is difficult to tell from this time-frame whether there is a real trend in temperature. More detailed analysis is required for that.

```{r fig.height = 7, fig.width = 8, echo=FALSE}
start <- bf.daily.tt$DATE_TIME[start(bf.daily.tt$DATE_TIME)[[1]]][[1]]
bf.mean.temp.ts <- ts(zoo::na.approx(bf.daily.tt$mean.temp), frequency = 365,start = start)

mean.temp.decomp <- decompose(bf.mean.temp.ts, type = "multiplicative")
par(mfrow=c(4,1))
plot(mean.temp.decomp$trend, main = "Mean Temperature Decomposition")
plot(mean.temp.decomp$seasonal)
plot(mean.temp.decomp$random)
plot(mean.temp.decomp$figure)
```

#### Mean Daily Wind Speed

The decomposition shows that there is likely yearly seasonality in the wind speed data. Once again, the random output shows likely seasonality as well. This indicates further work is required to finish the decomposition.

The trend retrieved from this decomposition is interesting. It shows that average daily wind speed is decreasing since 1973. 

```{r fig.height = 7, fig.width = 8, echo=FALSE}
start <- bf.daily.tt$DATE_TIME[start(bf.daily.tt$DATE_TIME)[[1]]][[1]]
bf.mean.spd.ts <- ts(zoo::na.approx(bf.daily.tt$mean.spd), frequency = 365,start = start)

mean.spd.decomp <- decompose(bf.mean.spd.ts, type = "additive")
par(mfrow=c(4,1))
plot(mean.spd.decomp$trend, main = "Mean Wind Speed Decomposition")
plot(mean.spd.decomp$seasonal)
plot(mean.spd.decomp$random)
plot(mean.spd.decomp$figure)
```

#### Daily Precipitation Decomposition

The decomposition of rainfall is fairly inconclusive. Visual review of the charts suggest that there may be higher daily rainfall possible during summer months (when there are thunderstorms and hurricanes), but the decomposition is not clean.

The data either needs to be transformed or a different method of decomposition needs to be used to validate whether the thunderstorm hypothesis is correct. 

For the needs of this analysis it would be better to simply compare rainfall to air quality data with and without lag.

```{r fig.height = 7, fig.width = 8, echo=FALSE}
start <- bf.daily.tt$DATE_TIME[start(bf.daily.tt$DATE_TIME)[[1]]][[1]]
bf.precip.ts <- ts(zoo::na.approx(filter_time(bf.daily.tt,'2001' ~ '2017')$sum.precip), frequency = 365,start = start)

precip.decomp <- decompose(bf.precip.ts, type = "multiplicative")
par(mfrow=c(4,1))
plot(precip.decomp$trend, main = "Daily Precipitation Decomposition")
plot(precip.decomp$seasonal)
plot(precip.decomp$random)
plot(precip.decomp$figure)
```

# **Conclusions**

As windspeed increases, Particulate Matter Decreases.

As temperature increases, probability of high PM2.5 increases.

Heavey rainfall coincides with low PM counts.

Forecast model was not accurate.

External factors outside of weather were not considered (such as the effects of regulation or industrial changes).